---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Comparison of community and citizen natural resource rights across Sub-Saharan Africa"
subtitle: "https://github.com/krm75/Environmental_Data_Analytics_2020/tree/master/ResourceRights"
author: "Kim Myers"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
setwd("../Data")

# Load your packages
library(ggplot2)
library(tidyverse)
library(knitr)
library(kableExtra)
library(wesanderson)
library(sf)
library(ggmap)
library(cowplot)
library(corrplot)
library(Hmisc)

# Set your ggplot theme
mytheme <- theme_light(base_size = 11) +
  theme(axis.text = element_text(color = "dark gray"), 
        legend.position = "right")

theme_set(mytheme)

# Load your datasets
wildlife <- read.csv("rights_to_resources_wildlife.csv")
water <- read.csv("rights_to_resources_water.csv")
trees <- read.csv("rights_to_resources_trees.csv")
petroleum <- read.csv("rights_to_resources_petroleum.csv")
minerals <- read.csv("rights_to_resources_minerals.csv")
africa <- st_read("Africa.shp")



```


# Rationale and Research Questions

Throughout much of the developing world, property and resources are controlled by the state. Conversely, these assets are most often privately owned in developed countries.  This difference has created vast inequalities in individual and community access to natural resources, which can be integral to livelihoods and health.  WRI's Rights to Resources dataset provides some insight into the state of natural resource rights in Sub-Saharan African countries, which are predominantly considered developing nations.  The data are formatted into 11 yes-no survey questions that specify limitations on resource access, with responses for 49 countries (Table 1; Table 2).  Using spatial and statistical analysis, I analyzed these survey data to find how resource rights vary by type and by country.  My research questions included:


## Question 1: How do individual and community rights in Sub-Saharan Africa differ between trees, wildlife, petroleum, water, and minerals?

## Question 2: How do individual and community rights differ between Sub-Saharan African countries?

By generating these results, I inferred which nations in the study region have the strongest and weakest policies supporting resource rights and for which resources.  Further research could link the resulting indicators of support to other variables, such as income and economic growth, to quantify the relationship between resource access and poverty.

```{r}
surveyqs <- c("Does the law provide that the natural resource is the property of the state or part of the public domain?",
              "Does the law recognize any private ownership of the natural resource?",
              "Does the law recognize customary or traditional rights to the natural resource?",
              "Does the law provide for the use of the natural resource by any person without authorization from the      government or payment of fees (hereafter called “free” natural resource use rights)?", 
              "By law, are “free” natural resource use rights linked to land rights or are they independent of land rights?",
              "By law, can a person employ any equipment, technology or method for exercising “free” natural resource use rights?",
              "By law, can “free” natural resource use rights be extended for commercial purposes without a government authorization?",
              "Does the law provide certain conditions, restrictions or limitations on the exercise of “free” natural resource use rights?",
              "Does the law provide for registration of “free” natural resource use rights? If not, does the law empower an institution to monitor the exercise of “free” use rights?", 
              "Does the law provide for revocation or taking away of “free” natural resource use rights?", 
              "Does the law have grievance and dispute resolution mechanisms available to holders of “free” natural resource use rights?")

qnums <- c('Q1','Q2','Q3','Q4','Q5','Q6','Q7','Q8','Q9','Q10','Q11')

survey.table <- data.frame(qnums,surveyqs)
colnames(survey.table) <- c('','Question')

kable(survey.table,
caption = "Table 1: List of survey questions included in the Rights to Resources dataset.", row.names = F)  %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),full_width = F)
```
```{r}
responses <- data.frame(c('No','No information available','Partial','Silent','Yes','Yes, implied'))

kable(Responses,
caption = "Table 2: Possible survey question responses.", row.names = F,
col.names = "Responses")  %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),full_width = F)

```




\newpage

# Dataset Information

The "Rights to Resource" data used in this analysis come from World Resource Institute's Land and Resource Rights Initiative and were published in November 2013.  In the data set, there are 5 sets of 11-question surveys detailing policies on water, trees, wildlife, minerals, and petroleum resources (Table 3).  Questions in the survey pertain to individual and community resource rights in sub-Saharan Africa.  In total, the project collected data on framework laws for 49 countries. Many of the nations studied enforce state-managed resource use. Resource Watch intended for this information to empower citizens and communities, investors, and policymakers to promote change. All data published by WRI was subjected to strict academic standards.  They strived to disperse information that is 'rigorous' and 'objective'.

To quantify support for resource rights, I excluded survey questions that did not exhibit a clear distinction between supporting and not supporting resource rights (Q5). If the response promoted the use of private natural resource use (could be affirmative or not affirmative, depending on the question), it was assigned a value of '1'.  If it did not, it received a value of '0'.  In this format, I summed the total number of supportive laws and divided by the total number of relevant questions to find the proportion of laws that promote resource rights.


```{r}
kable(head(wildlife),
caption = "Table 3: Sample of the first 10 rows of the 5 raw data sets.  There is one table for each resource.") %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),full_width = F) %>%
  column_spec(3:13, width_min = "4em", width_max = "4em")
```




\newpage

# Exploratory Analysis 

## Inspecting the data sets

Before wrangling the data, I looked at the structure of each data set.  I found that each resource table had 49 lines, one for each country surveyed.  I was able to list out all of the Sub-Saharan nations included in the analysis and the total numbers of columns - 13 - which comprised of all 11 survey questions and 2 fields for country ID.  Finally, I confirmed that all fields in the data were considered factors and viewed the range of responses to the surveys.

```{r}
length(unique(wildlife$Country))
```

```{r}
unique(wildlife$Country)
```

```{r}
nrow(wildlife)
```

```{r}
ncol(wildlife)
```

```{r}
colnames(wildlife)
```

```{r}
str(wildlife)
```
```{r}
summary(wildlife[,c(3:11)])
```
Looking at the range of factors for each survey question, I discovered that possible responses included: 'No information available', 'Partial', 'Yes', 'Yes, implied', 'No', and 'No, implied'.


##Quantifying the survey responses

In order to infer support for resource rights, I converted the survey responses to binary.  Any values in questions 2, 3, 4, 6, 7, 9, or 11 that equalled 'Yes', 'Yes, implied', and 'Partial' were assigned a value of 1.  I set all other responses ('No', 'No information available', No, implied', 'Silent') to 0. I repeated this process for questions 1, 8, and 10 but with a value of 1 for negative answers.  For this classification, I used only questions that clearly referred to the support for individual resource rights.  As a result, I included only 10 of the 11 survey questions in my analysis.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#summary(wildlife)
yes <- c('Yes','Yes, implied','Partial')
no <- c('No','No, implied','Silent')
unknown <- 'No information available'


wildlife[,c(4:6,8:9,11,13)] <- sapply(wildlife[,c(4:6,8:9,11,13)], FUN = function(x) ifelse(x %in% yes, 1, 0))
wildlife$county <- apply(wildlife[,c(4:6,8:9,11,13)],1,sum)

minerals[,c(4:6,8:9,11,13)] <- sapply(minerals[,c(4:6,8:9,11,13)], FUN = function(x) ifelse(x %in% yes, 1, 0))
minerals$county <- apply(minerals[,c(4:6,8:9,11,13)],1,sum)

water[,c(4:6,8:9,11,13)] <- sapply(water[,c(4:6,8:9,11,13)], FUN = function(x) ifelse(x %in% yes, 1, 0))
water$county <- apply(water[,c(4:6,8:9,11,13)],1,sum)

petroleum[,c(4:6,8:9,11,13)] <- sapply(petroleum[,c(4:6,8:9,11,13)], FUN = function(x) ifelse(x %in% yes, 1, 0))
petroleum$county <- apply(petroleum[,c(4:6,8:9,11,13)],1,sum)

trees[,c(4:6,8:9,11,13)] <- sapply(trees[,c(4:6,8:9,11,13)], FUN = function(x) ifelse(x %in% yes, 1, 0))
trees$county <- apply(trees[,c(4:6,8:9,11,13)],1,sum)



wildlife[,c(3,10,12)] <- sapply(wildlife[,c(3,10,12)], FUN = function(x) ifelse(x %in% no, 1, 0))
wildlife$countn <- apply(wildlife[,c(3,10,12)],1,sum)

minerals[,c(3,10,12)] <- sapply(minerals[,c(3,10,12)], FUN = function(x) ifelse(x %in% no, 1, 0))
minerals$countn <- apply(minerals[,c(3,10,12)],1,sum)

water[,c(3,10,12)] <- sapply(water[,c(3,10,12)], FUN = function(x) ifelse(x %in% no, 1, 0))
water$countn <- apply(water[,c(3,10,12)],1,sum)

petroleum[,c(3,10,12)] <- sapply(petroleum[,c(3,10,12)], FUN = function(x) ifelse(x %in% no, 1, 0))
petroleum$countn <- apply(petroleum[,c(3,10,12)],1,sum)

trees[,c(3,10,12)] <- sapply(trees[,c(3,10,12)], FUN = function(x) ifelse(x %in% no, 1, 0))
trees$countn <- apply(trees[,c(3,10,12)],1,sum)


wildlife$total.count <- wildlife$countn + wildlife$county
trees$total.count <- trees$countn + trees$county
minerals$total.count <- minerals$countn + minerals$county
water$total.count <- water$countn + water$county
petroleum$total.count <- petroleum$countn + petroleum$county


countrycounts <- data.frame(wildlife$Country,wildlife$total.count,minerals$total.count,water$total.count,petroleum$total.count,trees$total.count)
countrycounts$Total <- apply(countrycounts[,c(2:6)],1,sum)/50
colnames(countrycounts) <- c("Country","Wildlife","Minerals","Water","Petroleum","Trees","Total")

kable(countrycounts[order(-countrycounts$Total),],
caption = "Table 4: Number of survey question responses (n=10) that support individual/community resource rights to some degree.", row.names = F)  %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),full_width = F) #%>%
  #column_spec(3:13, width_min = "4em", width_max = "4em")

```

Table 4 illustrates the total number of survey questions that support individual or community resource rights for each resource type in each country.  Through these calculations, I was able to quantify the proportion of supportive responses for each country studied.


\newpage

# Analysis

Insert visualizations and text describing your main analyses. Format your R chunks so that graphs are displayed but code and other output is not displayed. Instead, describe the results of any statistical tests in the main text (e.g., "Variable x was significantly different among y groups (ANOVA; df = 300, F = 5.55, p < 0.0001)"). Each paragraph, accompanied by one or more visualizations, should describe the major findings and how they relate to the question and hypotheses. Divide this section into subsections, one for each research question.

```{r fig1, echo=FALSE,message=FALSE,include=TRUE,fig.cap="Average proportion of survey responses which support individual/community resource rights to some degree by resource type."}
country.counts.rcorr <- cor(countrycounts[,c(2:4,6)])

corrplot(country.counts.rcorr) #discluded petroleum

```

```{r fig1, echo=FALSE,message=FALSE,include=TRUE,fig.cap="Average proportion of survey responses which support individual/community resource rights to some degree by resource type."}

average.yeses <- apply(countrycounts[,c(2:6)],2,mean)/7 #CHECK THIS
stddev.yeses <- (apply(countrycounts[,c(2:6)],2,sd)/7)/2
resources <- c('Wildlife','Minerals','Water','Petroleum','Trees')
yeses.figure.df <- data.frame(resources,average.yeses,stddev.yeses,row.names = NULL)
colnames(yeses.figure.df) <- c('resource','average','stddev')
yeses.figure.df <- yeses.figure.df[order(-yeses.figure.df$average),]

average.yeses.plot <- ggplot(yeses.figure.df, aes(x=reorder(resource, -average), y=average, fill=resource)) + 
  geom_bar(stat="identity",position=position_dodge()) +
  geom_errorbar(aes(ymin=average-stddev, ymax=average+stddev), width=.2,
               position=position_dodge(.9)) +
  labs(x="", y = "Average Porportion of 'Yes' Responses",fill="") +
  scale_fill_manual(values=wes_palette(name="Cavalcanti1",n=5))
print(average.yeses.plot)

```



```{r,echo=FALSE,message=FALSE,include=TRUE,fig.cap="Total number of survey questions from the Rights to Resources dataset that support individual or community resource rights.  The highest possible sum for each resource was 7.  In the All Resources panel, values equal the proportion of responses that promote resource rights."}

#attributes(africa_wildlife_merged) 
africa_resources_merged <- merge(africa, countrycounts, by.x = "COUNTRY", by.y = "Country", all=T)

africa_total <- ggplot(africa_resources_merged) +
  geom_sf(aes(fill = Total),col='black') +
  labs(title='All Resources',fill='') +
  scale_fill_gradient(low="white", high="#4E2A1E",na.value = 'gray')

africa_wildlife <- ggplot(africa_resources_merged) +
  geom_sf(aes(fill = Wildlife),col='black') +
  labs(title='Wildlife',fill='') +
  scale_fill_gradient(low="white", high="#972D15",na.value = 'gray')

africa_trees <- ggplot(africa_resources_merged) +
  geom_sf(aes(fill = Trees),col='black') +
  labs(title='Trees',fill='')+
  scale_fill_gradient(low="white", high="#A2A475",na.value = 'gray')

africa_water <- ggplot(africa_resources_merged) +
  geom_sf(aes(fill = Water),col='black') +
  labs(title='Water',fill='')+
  scale_fill_gradient(low="white", high="#81A88D",na.value = 'gray')

africa_minerals <- ggplot(africa_resources_merged) +
  geom_sf(aes(fill = Minerals),col='black') +
  labs(title='Minerals',fill='')+
  scale_fill_gradient(low="white", high="#FAD510",na.value = 'gray')

africa_petroleum <- ggplot(africa_resources_merged) +
  geom_sf(aes(fill = Petroleum),col='black') +
  labs(title='Petroleum', fill='')+
  scale_fill_gradient(low="white", high="white",na.value = 'gray')


plot_grid(africa_total,africa_wildlife,africa_trees,africa_water,africa_minerals,africa_petroleum, nrow = 2, align = 'h', rel_heights = c(2, 2))

```




 




\newpage

# Summary and Conclusions

Summarize your major findings from your analyses in a few paragraphs. What conclusions do you draw from your findings? Relate your findings back to the original research questions and rationale.
